---
title: "Final_Report"
author: "Arely Salinas Cambron, Ruby Cruz, Kylize Dela Cruz, and Raquel Gonzalez"
date: "2026-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Day 1: Environmental Setup & Data Import
## Task 2: Environment Setup & Data Setup

```{r}
# Load libraries
library(tidyverse)
library(psych)
```

```{r}
# Import dataset using relative path for reproducibility
nhis_data <- read.csv("data/NHIS _Data_2021.csv")
```

```{r}
# Initial exploration
str(nhis_data)
summary(nhis_data)
head(nhis_data)
```
### Interpretation

The initial exploration of the NHS 2021 dataset reveals a robust sample of 29,482 participants. However, the summary() output highlights the critical necessity of the data cleaning steps performed in the next task. Specifically, variables such as weight and height show extreme maximum values like 999 and 99, which correspond to non-response codes rather than biological measurements. Further, the categorical variables contain placeholder integers for missing data that would invalidate frequency tables if left unaddressed. This exploration justifies the transition to a cleaned dataset to ensure the integrity of our public health analysis. 

# Day 2: Data Cleaning & Management
## Task 1: Variable Selection & Missing Value Check

```{r}
# Select only the required variables
nhis_subset <- nhis_data %>%
  select(AGEP_A, WEIGHTLBTC_A, HEIGHTTC_A, SEX_A, HISPALLP_A, EDUCP_A, PHSTAT_A, LSATIS4R_A)
```

```{r}
# Removing missing and non-response codes by removing rows where any columns have codes 7, 8, 9, 97, 98, 99, 996, 997, 998, or 999
# Nest ifelse() with respective codes into mutate()
# Change codes to NAs, then delete at the end
nhis_clean <- nhis_subset %>%
  # Remove 97-99 from AGEP_A
  mutate(AGEP_A = ifelse(AGEP_A >= 97, NA, AGEP_A)) %>%
  # Remove 996-999 from WEIGHTLBTC_A
  mutate(WEIGHTLBTC_A = ifelse(WEIGHTLBTC_A >= 996, NA, WEIGHTLBTC_A)) %>%
  # Remove 96-99 from HEIGHTTC_A
  mutate(HEIGHTTC_A = ifelse(HEIGHTTC_A >= 96, NA, HEIGHTTC_A)) %>%
  # Remove 97-99 from HISPALLP_A and EDUC_A
  # Use across() to work with multiple variables at once
  mutate(across(c(HISPALLP_A, EDUCP_A),
                ~ ifelse(.x %in% c(97, 98, 99), NA, .x))) %>%
  # Remove 7, 8, 9 from SEX_A, PHSTAT_A, and LSATIS4R_A
  # Use across() to work with multiple variables at once
  mutate(across(c(SEX_A, PHSTAT_A, LSATIS4R_A),
                ~ ifelse(.x %in% c(7, 8, 9), NA, .x))) %>%
  # Remove all rows that now have an NA
  drop_na()
```

### Interpretation

The data cleaning process was essential to transform the raw NHIS dataset into a reliable analytical sample for public health research. The first step involved a targeted variable selection, narrowing the scope of the data to eight key determinants, including age, weight, height, and various sociodemographic factors. The first cleaning step included identifying and converting non-response codes (based on the NHIS codebook) into NA values. This prevents placeholder values for "Refused" or "Don't Know" from being treated as actual measurements, which would have severely biased the physiological averages, such as incorrectly inflating the mean weight of the cohort.

Once the missing values were standardized, listwise deletion was applied to ensure a complete-case analysis. While this reduces the overall sample size, it is a necessary choice in this context to ensure that every participant included in the final analysis has a full set of data across all variables. This provides a consistent baseline for calculating correlations and health trends. 

## Task 2: Recoding Variables

```{r}
# Update nhis_clean by recoding EDUC_A variable based on codebook categories
nhis_clean <- nhis_clean %>%
  # Use case_when() instead of factor() to first group numbers into 4 categories first
  # Save to new variable Education to save original EDUCP_A data
  mutate(Education = case_when(
    EDUCP_A >= 0 & EDUCP_A <= 3 ~ "less than High School",
    EDUCP_A == 4 ~ "High School Graduate",
    EDUCP_A >= 5 & EDUCP_A <= 7 ~ "Some College Education",
    EDUCP_A >= 8 & EDUCP_A <= 10 ~ "College Graduate or better")) %>%
  # Change category names into a factor with specific labels and a specific order
  mutate(Education = factor(Education, levels = c("less than High School", "High School Graduate", "Some College Education", "College Graduate or better")))
```

```{r}
# Save the final cleaned dataset as a csv in the data folder
write_csv(nhis_clean, "data/nhis_clean.csv")
```

### Interpretation

To make the data more actionable for public health interpretation, the education variable (EDUCP_A) was recoded from its original eleven categories into four broader educational tiers. This "collapsing" of categories is a standard practice in epidemiological research to increase statistical power and create more stable groupings for comparison. By grouping education levels into categories such as "less than High School" or "College Graduate or better", this variable serves as a more effective proxy for socioeconomic status. This allows for clearer analysis of how educational attainment intersects with behavioral health attributes and socioeconomic status. 

This recoded variable was converted into a factor with meaningful level names. This transition from raw numeric codes to descriptive labels ensures that the resulting visualizations and summary tables are easily interpretable for public health stakeholders. By explicitly ordered the factor levels, the analysis preserves the natural hierarchy of education, ensuring that any identified trends, such as the correlation between higher education and improved life satisfaction, are presented in a logical and professional manner. This refined dataset, saved as nhis_clean.csv, serves as the finalized foundation for all univariate and bivariate reporting.

# Day 3: Univariate & Bivariate Analysis & Visualization

## Task 1: Univariate Analysis

The univariate analysis provides a foundational understanding of the individual characteristics within our cleaned NHIS 2021 dataset. By examining quantitative and qualitative variables separately, we can describe the central tendencies, spread, and frequency distributions of the participants. 

### Quantitative Variables

```{r}
# Summary statistics for quantitative variables AGEP_A, WEIGHTLBTC_A, HEIGHTC_A

# Age
cat("\nAGEP_A Summary:\n")
mean(nhis_clean$AGEP_A, na.rm = TRUE)
median(nhis_clean$AGEP_A, na.rm = TRUE)
sd(nhis_clean$AGEP_A, na.rm = TRUE)
summary(nhis_clean$AGEP_A)
```

```{r}
# Weight
cat("\nWEIGHTLBTC_A Summary:\n")
mean(nhis_clean$WEIGHTLBTC_A, na.rm = TRUE)
median(nhis_clean$WEIGHTLBTC_A, na.rm = TRUE)
sd(nhis_clean$WEIGHTLBTC_A, na.rm = TRUE)
summary(nhis_clean$WEIGHTLBTC_A)
```

```{r}
# Height
cat("\nHEIGHTTC_A Summary:\n")
mean(nhis_clean$HEIGHTTC_A, na.rm = TRUE)
median(nhis_clean$HEIGHTTC_A, na.rm = TRUE)
sd(nhis_clean$HEIGHTTC_A, na.rm = TRUE)
summary(nhis_clean$HEIGHTTC_A)
```

```{r}
# Base R Histogram

hist(nhis_clean$AGEP_A,
     main = "Histogram of Age",
     xlab = "Age",
     col = "blue")
# Histogram is slightly right skewed and indicates most participants are 50-65 years old. 

hist(nhis_clean$WEIGHTLBTC_A,
     main = "Histogram of Weight",
     xlab = "Weight",
     col = "blue")
# Most participants are 140-190 pounds 

hist(nhis_clean$HEIGHTTC_A,
     main = "Histogram of Height",
     xlab = "Height",
     col = "blue")
# Participants are mostly around 64-69 inches 

```

```{r}
# Ggplot2 histogram

ggplot(nhis_clean, aes(x = AGEP_A)) +
  geom_histogram(bins = 30, fill = "lightblue") +
  ggtitle("Age Distribution")
# Participants age is spread across 20-80 years old 

ggplot(nhis_clean, aes(x = WEIGHTLBTC_A)) +
  geom_histogram(bins = 30, fill = "lightblue") +
  ggtitle("Weight Distribution")
# Most participants are 140-200 pounds, but participants go up to 300 pounds

ggplot(nhis_clean, aes(x = HEIGHTTC_A)) +
  geom_histogram(bins = 30, fill = "lightblue") +
  ggtitle("Height Distribution")
# Most participants are around 65-70 inches
```

```{r}
# Base R Boxplot

boxplot(nhis_clean$AGEP_A,
        main = "Age Boxplot",
        col = "green")
# Average age of participants is around 55 years old

boxplot(nhis_clean$WEIGHTLBTC_A,
        main = "Weight Boxplot",
        col = "green")
# Average weight of participants is about 175 pounds

boxplot(nhis_clean$HEIGHTTC_A,
        main = "Height Boxplot",
        col = "green")
# Average height of participants are about 66 inches
```

```{r}
# ggplot2 boxplots

ggplot(nhis_clean, aes(y = AGEP_A)) +
  geom_boxplot(fill = "lightgreen") +
  ggtitle("Age Boxplot")
# Average age of participants is 54 years old

ggplot(nhis_clean, aes(y = WEIGHTLBTC_A)) +
  geom_boxplot(fill = "lightgreen") +
  ggtitle("Weight Boxplot")
# Average weight of participants is 174 lbs

ggplot(nhis_clean, aes(y = HEIGHTTC_A)) +
  geom_boxplot(fill = "lightgreen") +
  ggtitle("Height Boxplot")
# Average participants height is 66 inches 
```

### Interpretation

The summary statistics and visualizations for the quantitative variables reveal a diverse adult population with expected physiological ranges.

Age(AGEP_A): The average age of the sample is approximately 52.6 years, with a median of 54, suggesting a relatively balanced distribution of middle-aged adults. The histograms show a slight right skew, indicating a wide representation across the lifespan from 18 to 85 years. The boxplots confirm that 50% of the participants fall between the ages of 37 and 67, the IQR.

Weight(WEIGHTLBTC_A): Participants' weights range from 100 to 299 pounds, with a mean weight of 176.8 pounds. Both the histograms and boxplots illustrate that the majority of the population is concentrated between 140 and 200 pounds. The boxplot highlights that while the distribution is centered around the median of 173 lbs, there is a notable spread toward the upper weight limit.

Height(HEIGHTTC_A): The height distribution is approximately normal, centered around a mean of 66.7 inches, or roughly 5'7". The narrow standard deviation at 3.9 inches and the boxplot indicate that most participants are between 64 and 70 inches tall.

### Qualitative Variables

```{r}
# Frequency tables and graphs for qualitative variables

# 1. SEX_A

table(nhis_clean$SEX_A)

barplot(table(nhis_clean$SEX_A),
        main = "Sex Distribution",
        xlab = "Sex",
        ylab = "Frequency",
        col = "lightblue")

ggplot(nhis_clean, aes(x = factor(SEX_A))) +
  geom_bar(fill = "blue") +
  labs(title = "Sex Distribution",
       x = "Sex",
       y = "Frequency") +
  theme_minimal()
```

```{r}
# 2. HISPALLP_A

table(nhis_clean$HISPALLP_A)

barplot(table(nhis_clean$HISPALLP_A),
        main = "Hispanic Origin Distribution",
        xlab = "Hispanic Origin",
        ylab = "Frequency",
        col = "lightgreen")

ggplot(nhis_clean, aes(x = factor(HISPALLP_A))) +
  geom_bar(fill = "green") +
  labs(title = "Hispanic Origin Distribution",
       x = "Hispanic Origin",
       y = "Frequency") +
  theme_minimal()

```

```{r}
# 3. EDUCP_A
table(nhis_clean$Education)

barplot(table(nhis_clean$Education),
        main = "Education Level Distribution",
        xlab = "Education Level",
        ylab = "Frequency",
        col = "orange")

ggplot(nhis_clean, aes(x = Education)) +
  geom_bar(fill = "orange") +
  labs(title = "Education Level Distribution",
       x = "Education Level",
       y = "Frequency") +
  theme_minimal()
```

```{r}
# 4. PHSTAT_A
table(nhis_clean$PHSTAT_A)

barplot(table(nhis_clean$PHSTAT_A),
        main = "Physical Health Status Distribution",
        xlab = "Physical Health Status",
        ylab = "Frequency",
        col = "purple")

ggplot(nhis_clean, aes(x = factor(PHSTAT_A))) +
  geom_bar(fill = "purple") +
  labs(title = "Physical Health Status Distribution",
       x = "Physical Health Status",
       y = "Frequency") +
  theme_minimal()
```

```{r}
# 5. LSATIS4R_A
table(nhis_clean$LSATIS4R_A)

barplot(table(nhis_clean$LSATIS4R_A),
        main = "Life Satisfaction Distribution",
        xlab = "Life Satisfaction",
        ylab = "Frequency",
        col = "pink")

ggplot(nhis_clean, aes(x = factor(LSATIS4R_A))) +
  geom_bar(fill = "pink") +
  labs(title = "Life Satisfaction Distribution",
       x = "Life Satisfaction",
       y = "Frequency") +
  theme_minimal()
```

### Interpretation 

Frequency tables and bar plots were used to assess the distribution of categorical variables. These distributions provide context for the social and health-related tiers of the sample.

Sex(SEX_A): The frequency table and bar plots indicate that Females ("2") make up a slight majority of the sample, providing a relatively even basis for gender-based health comparisons.

Hispanic Origin(HISPALLP_A): The bar chart reveals that a large majority of participants identify as Non-Hispanic White (n = 17,617), followed by Hispanic (n =3,533). This distribution is crucial for identifying potential health disparities across different racial and ethnic groups in the subsequent analysis.

Education Level(Education): Based on the recoded variable, the largest segment of our population consists of "College Graduates or better". This highlights the socioeconomic diversity of the sample and serves as a primary independent variable for analyzing health status.

Physical Health Status(PHSTAT_A): The distribution of general health status is skewed toward the "Very Good" (2) categories, followed by "Good", and then "Excellent." This suggests that a significant portion of the cohort perceives their physical health as positive, though a smaller frequency remains in the "Fair" and "Poor" categories.

Life Satisfaction(LSATIS4R_A): Life satisfaction plots show that most participants reported being "Very Satisfied" (1) and "Satisfied" (2). Understanding the frequency of these subjective well-being scores is essential for exploring the relationship between mental outlook and physical health outcomes.  

The univariate analysis confirms that the cleaned dataset is free from the extreme outliers identified during the inital exploration. The cohort is primarily composed of middle-aged adults with a broad range of educational backgroynds and generally positive self-reported health and life satisfaction. These distributions provide the necessary context to move forward with bivariate analysis to see how these variables interact.

## Task 2: Bivariate Analysis

The bivariate analysis explores the relationships between pairs of variables to identify patterns, differences, and correlations within the NHIS 2021 cohort. By examining how quantitative measures vary across qualitative categories, and how two quantitative measures interact, we gain deeper insight into the health profile of this sample.

### Quantitative vs Qualitative Boxplots

```{r}
# Creating boxplots for AGEP_A and to display levels by SEX_A and EDUCP_A

boxplot(AGEP_A ~ factor(SEX_A), data = nhis_clean,
        main="Boxplot of Age by Sex",
        xlab = "Sex",
        ylab= "Age",
        col= "skyblue", 
        border="black")

boxplot(AGEP_A ~ Education, data = nhis_clean,
        main= "Boxplot of Age by Education Level",
        xlab = "Education",
        ylab = "Age",
        col= "skyblue",
        border= "black")

ggplot(nhis_clean, aes(x=SEX_A, y=AGEP_A))+
  geom_boxplot(aes(fill = factor(SEX_A)))+
  labs(title = "Boxplot of Age by Sex",
       x= "Sex",
       y= "Age")+
  theme_minimal()

ggplot(nhis_clean, aes(x=Education, y=AGEP_A))+
  geom_boxplot(aes(fill = factor(Education)))+
  labs(title = "Boxplot of Age by Education Level",
       x= "Education Level",
       y= "Age")+
  theme_minimal()
```

### Quantitative vs Qualitative Clustered Bar Chart

```{r}
# Creating clustered bar chart for PHSTAT_A (General Health) by LSATIS4R_A (Life Satisfaction)

ggplot(nhis_clean, aes(x=factor(LSATIS4R_A), fill= factor(PHSTAT_A)))+
  geom_bar(position = "dodge")+
  labs(title = "Clustered Bar Chart of General Health Status by Quality of life",
       x="Life Satisfaction",
       y="General Health Status")+
    theme_minimal()
```

### Quantitative vs Qualitative Scatter Plots

```{r}
# Creating a scatter plot for HEIGHTTC_A and WEIGHTLBTC_A

x <- nhis_clean$HEIGHTTC_A
y <- nhis_clean$WEIGHTLBTC_A

plot(x,y,
     main="Scatter plot of Height vs Weight",
     xlab = "Height",
     ylab= "Weight", 
     pch = 19, frame = FALSE)

ggplot(nhis_clean,
       aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A))+
  geom_point(color="purple",
             size= 2,
             alpha=.6)+
  labs(title = "Scatter plot of Height vs Weight",
       x= "Height",
       y= "Weight")

# Correlation report Height vs Weight
# This report shows that there is a significant relationship p-value<2.2e-16, because we reject the null hypothesis.

cor.test(nhis_clean$HEIGHTTC_A,nhis_clean$WEIGHTLBTC_A)
```

### Interpretation

Side-by-side boxplots were utilized to visualize how age is distributed across different demographic groups. The distribution of age appears similar between males and females. The median age for both groups remains near the mid-50's, suggesting that gender does not significantly skew the age profile of this specific cohort. The boxplots also visualize age across the four recoded educational levels: less than High School, High School Graduate, Some College Education, and College Graduate or better. The "less than High School" group appears to have a slightly higher median age, which likely reflects broader historical shifts in education access, where older generations were less likely to obtain high school diplomas or college degrees compared to younger cohorts. 

The clustered bar chart was employed to examine the relationship between a participant's perceived physical health status (PHSTAT_A) and their reported life satisfaction (LSATIS4R_A). The visualization shows a clear positive association between physical health and mental well-being. Participants who reported being "Very Satisfied (1) were predominantly represented in the "Excellent" (1) and "Very Good" (2) health categories, with very few reporting poor health. Lower levels of satisfaction shifts the modal health response towards "Good" (3), "Fair" (4), and "Poor" (5). This suggests that high life satisfaction may act as a protective factor or a marker for overall physical resilience. The dissatisfied individuals highlight a vulnerable population that may face a double burden of both physical ailment and lower psychological well-being.

To assess the relationship between physical measurements, a scatter plot and a Pearson correlation test were conducted for height (HEIGHTTC_A) and weight (WEIGHTLBTC_A). The scatter plot demonstrates a clear positive linear trend, indicating that as height increases, weight generally tends to increase as well. The vertical strips in the plot reflect that height was recorded in whole-inch increments, ranging from 59 to 76 inches. The correlation analysis yielded a p-value of < 2.2 * 10^-16, which is far below the standard significance threshold of 0.05. This highly significant result allows us to reject the null hypothesis and confirm a genuine relationship between these variables in the NHIS population. The calculated Pearson's r is 0.502, indicating a moderate-to-strong positive correlation. This means that approximately 25% (r^2 = 0.25) of the variation in weight can be explained by height alone.

The bivariate results highlight critical associations. While age remains stable across sex, it varies by educational attainment due to generational trends. Most notable, the data confirms a strong link between physical health and life satisfaction, alongside a highly significant correlation (r = 0.502) between height and weight. These findings provide the groundwork for multivariate analysis, where we will see if these patterns hold when considering multiple factors simultaneously.

# Day 4: Multivariate Analysis

## Enhanced Scatterplot

```{r}
# Use ggplot2 to enhance the Day 3 Scatterplot
ggplot(nhis_clean, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A, color = factor(SEX_A))) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ Education) +
  labs(title = "Height vs Weight by Sex, Faceted by Education Level",
       x = "Height (inches)",
       y = "Weight (pounds)",
       color = "Sex (1=M, 2=F") +
  theme_minimal()
```

### Correlation Scatter Plot Matrix

```{r}

# Use pairs.panels() to create a correlation matrix
# Select only the quantitative variables for the matrix
quant_vars <- nhis_clean %>%
  select(AGEP_A, WEIGHTLBTC_A, HEIGHTTC_A)

pairs.panels(quant_vars,
             method = "pearson",
             hist.col = "skyblue",
             density = TRUE,
             ellipses = TRUE)
```

### Interpretation